Reasonably Programmable Literal Notation

### 2018-05-04 icfp18/rebuttal.txt

We thank the referees for their thoughtful reviews!

# Review A

Racket certainly does offer granular escape hatches out of its hygiene 
discipline. The problem is that the client does not know, abstractly, whether these 
escape hatches have been used strictly to support splicing from 
a string literal body.

To address this problem, it would be possible to use the escape hatches in Racket to implement, as a layer atop the Racket macro system, the literal macro system described in this paper. This point should also address the critique that the paper is "very closely tied to ML" --
in fact, the contributions are relevant even in dynamic languages like Racket 
(by simply ignoring the typing aspects, or by inserting a corresponding 
contract on each expansion). We will make these points clear in the final related work section.

With the additional page allowance in the final paper, we can include the HTML example from the supplement. It should not be necessary to remove the metatheory, which 
is a valuable contribution in that it makes the claimed reasoning 
principles precise (Theorems 5.2 and 5.3).

We would also be happy to provide the additional small clarifications suggested 
in the "Details" section of the review.

# Review B

> it does not appear that TLMs allow to easily splice in an arbitrary experssion of the host language

Spliced expressions can be any arbitrary expression of the host language that 
does not contain unbalanced \`( and )\`. Sec. 2.1 suggests this and many of the examples demonstrate this, but we would be happy to state this more directly.

The SQL example can certainly be expressed. If the "WHERE" expression appears directly in the expansion (i.e. it has a type like `DBRow -> bool`), this is straightforward. 
Alternatively, the TLM might internally extract its parse tree (like the 
`$re_expr` TLM in 2.2.2) and transform it. Any free 
variables in the transformed expression would need to be either 
(1) marked as spliced expressions, or (2) appear in the TLM's dependencies.

> I would have liked more details about the implementation

We would be happy to include the requested implementation details 
in the final version.

> would it be possible to rewrite the
second branch (with proper syntax support) :
```
| `( lam x . e )` as l => $v l
```

This is not possible -- the meaning of variables in OCaml is unchanged, 
so `l` stands for the matched value, not the generalized literal form 
appearing in the pattern. It is possible, of course, to instead apply a 
partial function `term_to_val : Lambda.term -> Lambda.v`.

# Review C

> the resulting system is presented only through a few partially-omitted 
examples

Fig. 3 together with Fig. 5 actually presents a complete example with no omissions, including the full 
lexer and parser. We would be happy to also include the second complete 
example from the supplement (see above).

There are in total 8 examples in the paper ($kq, $regex, $html, $smiles, 
$listcomp, $re_expr, \$term/\$v, $map). These span a variety
of application domains and demonstrate all of the features of the 
proposed system.

### 2018-06-12 icfp18/syntax-icfp18.tex
final submission
deanonymized + changed bib format + small edits to abstract

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
% \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
% \citestyle{acmauthoryear}   %% For author/year citations

@@

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}
}
\email{first2.last2@inst2b.org}         %% \email is recommended

@@

Compared to existing approaches, TLMs are uniquely \emph{reasonable}. TLM clients can reason abstractly, i.e. without examining grammars or generated expansions, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal body, which gives the locations and types of spliced subterms. TLM providers can reason modularly about syntactic ambiguity and expansion correctness according to clear criteria. This paper integrates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system. We discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness. Finally, we establish these abstract reasoning principles formally with a detailed type-theoretic account of expression and pattern TLMs for ``core ML''. 

@@

\begin{figure}[h]
\vspace{-9px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-4px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-7px}
\end{figure}

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
\documentclass[acmsmall]{acmart}
% \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[ICFP'18]{International Conference on Functional Programming}{September 23--29, 2018}{St. Louis, MO, USA}
\acmYear{2018}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

@@

%% Author with single affiliation.
\author{Cyrus Omar}
\authornote{The majority of this research was performed while the first author was at Carnegie Mellon University.}          %% \authornote is optional;
                                        %% can be repeated if necessary
\affiliation{
  \institution{University of Chicago}            %% \institution is required
  \city{Chicago}
  \state{IL}
  % \country{USA}
}
\email{comar@cs.uchicago.edu}          %% \email is recommended
\affiliation{
  \institution{Carnegie Mellon University}           %% \institution is required
  \city{Pittsburgh}
  \state{PA}
  % \country{USA}
}

%% Author with two affiliations and emails.
\author{Jonathan Aldrich}
\affiliation{
  \institution{Carnegie Mellon University}           %% \institution is required
  \city{Pittsburgh}
  \state{PA}
  % \country{USA}
}
\email{jonathan.aldrich@cs.cmu.edu}         %% \email is recommended

@@

Compared to existing approaches, TLMs are uniquely \emph{reasonable}. TLM clients can reason abstractly, i.e. without examining grammars or generated expansions, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal body, which gives the locations and types of spliced subterms. TLM providers can reason modularly about syntactic ambiguity and expansion correctness according to clear criteria. This paper incorporates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system. We discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness. Finally, we establish these abstract reasoning principles formally with a detailed type-theoretic account of expression and pattern TLMs for ``core ML''. 

@@

\begin{figure}[h]
\vspace{-9px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-4px}
\caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-7px}
\end{figure}
